---
title: "Exercise 3.2.4"
output: html_notebook
---

Load the `tidyverse` library

```{r}
library(tidyverse)
```


## Question 1: Run `ggplot(data = mpg)`. What do you see?

```{r}
library(tidyverse)
ggplot(data=mpg)
```

Answer: No graph


## Question 2: How many rows are in mtcars? How many columns?
### Rows:
Use `nrow` to count the number of rows in a data frame.
```{r}
mtcars_nrow <- nrow(mtcars)
mtcars_nrow
```
The answer is `r mtcars_nrow` rows.

### Columns:
Use `ncol` to count the number of columns in a data frame.

```{r}
mtcars_ncol <- ncol(mtcars)
mtcars_ncol
```

The answer is `r mtcars_ncol` rows.


## Question 3: What does the drv variable describe? Read the help for ?mpg to find out.
```{r}
?mpg

# f = front-wheel drive, r = rear wheel drive, 4 = 4wd
```

## Question 4: Make a scatterplot of hwy versus cyl.
```{r}
ggplot(mpg, aes(x = cyl, y = hwy)) + 
  geom_point()

#also
ggplot(mpg, aes(x = cyl, y = hwy)) + 
  geom_point()
```

## Question 5: What happens if you make a scatterplot of class versus drv? Why is the plot not useful?

```{r}
ggplot(mpg, aes(x = drv, y = class)) + 
  geom_point()

#also
ggplot(mpg, aes(x = class, y = drv)) + 
  geom_point()
```

This scatter plot may not be useful because the dots on the graph merely represent the availability of data within each combination of categories.  